---
title: "pmdplyr: Panel Maneuvers in dplyr"
author: "Nick Huntington-Klein, Philip Khor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
<!-- output: rmarkdown::html_vignette. pdf_document -->
vignette: >
  %\VignetteIndexEntry{pmdplyr: Panel Maneuvers in dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

```{r setup}
library(pmdplyr)
```

The `pmdplyr` package is an extension to `dplyr` designed for cleaning and managing panel and hierarchical data. It contains variations on the `dplyr::mutate()` and `dplyr::join()` functions that address common panel data needs, and contains functions for managing and cleaning panel data. 

Unlike other panel data packages, functions in `pmdplyr` are all designed to work even if there is more than one observation per individual per period. This comes in handy if each individual is observed multiple times per period - for example, multiple classes per student per term; or if you have hierarchical data - for example, multiple companies per country.

There are three vignettes in total describing the contents of `pmdplyr`: 

1. **CURRENT VIGNETTE** "pmdplyr"/"Get Started", which describes the `pibble` panel data object type, and the `pmdplyr` tools for creating well-behaved ID and time variables `id_variable()` and `time_variable()`.

2. "dplyr variants", which that describes `pmdplyr` variations on `dplyr` functions `mutate()` (`mutate_cascade()` and `mutate_subset()`), `_join` (`inexact_join` and `safe_join()`), and `lag` (`tlag()`).

3. "Panel Tools", which describes novel tools that `pmdplyr` provides for cleaning andd manipulating panel data (`panel_fill()`, `panel_locf()`, `fixed_check()`, `fixed_force()`, `between_i()`, `within_i()`, `mode_order()`).

-----

# pibble

The `pibble` data type is a `tibble` data frame with additional attributes `.i`, which is a set of variables in the `pibble` that identifies individuals, `.t`, which is a variable in the `pibble` that indentifies the time index, and `.d`, which identifies the gap between periods of `.t`. If the gap between time periods doesn't matter and any consecutive time periods should be treated as consecutive, set `.d = 0`.

`pibble` status will be maintained if the `pibble` is modified using functions from `pmdplyr` or `dplyr` (unless you use those functions in a way that drops or renames the `.i` or `.t` variables, in which case `pibble` status will be lost). Other data manipulation functions may remove `pibble` status, requiring it to be re-declared as a `pibble`. 

Most functions in `pmdplyr` will allow you to declare `.i` and `.t` in the function itself. But if a `pibble` is passed in via a `dplyr` verb, there is no need.

## pibble() and as_pibble()

`pibble`s can be declared in two main ways: raw, via `pibble()`:

```{r, eval=FALSE}
pibble(...,
  .i = NULL,
  .t = NULL,
  .d = 1,
  .uniqcheck = FALSE
)
```

or by transforming an existing `data.frame`, `list`, or `tbl_df` using `as_pibble()`:

```{r, eval=FALSE}
as_pibble(x,
  .i = NULL,
  .t = NULL,
  .d = 1,
  .uniqcheck = FALSE,
  ...
)
```

Both functions work exactly as `tibble::tibble()` and `tibble::as_tibble()` do, except that they also take the arguments `.i`, `.t`, and `.d`, with `.i` and `.t` accepting either unquoted or quoted variable names. If you'd like your `pibble` checked to see if `.i` and `.t` uniquely identify your observations, set `.uniqcheck = TRUE`. It will do this automatically the first time in each R session you create a `pibble`, but if you'd like it to keep doing it, use `uniqcheck`. 

As a side bonus, you can check if the variables `a, b, c` uniquely identify the observations in data set `d` by running `as_pibble(d, .i = c(a, b, c), .uniqcheck = TRUE)`. No warning? It's uniquely identified!

```{r}
# .d = 1 by default, so in this data,
# a = 1, b = 3 comes one period after a = 1, b = 2.
basic_pibble <- pibble(
  a = c(1, 1, 1, 2, 2, 2),
  b = c(1, 2, 3, 2, 3, 3),
  c = 1:6,
  .i = a,
  .t = b
)

data(SPrail)
# In SPrail, insert_date does not imply regular gaps between
# time periods, so we set .d = 0
declared_pibble <- as_pibble(SPrail,
  .i = c(origin, destination),
  .t = insert_date,
  .d = 0
)
```

## panel_convert()

`pmdplyr` also has the function `panel_convert()` which allows you to convert between different popular R panel data objects, including `pibble`. This can come in handy for creating `pibbles`, or exporting your cleaned `pibble` to use with a package that does panel data *analysis* (which `pmdplyr` does not):

```{r, eval = FALSE}
panel_convert(
  data,
  to,
  ...
)
```

Where `data` is a panel data object, either `pibble`, `tsibble`, `pdata.frame`, or `panel_data`, and `to` is the type of object you'd like returned, which you can refer to by object name, object class, or package name: get a `pibble` with `"pmdplyr"`, `"pibble"`, or `"tbl_pb"`, a `tsibble` with `"tsibble"` or `"tbl_ts"`, a `pdata.frame` with `"plm"` or `"pdata.frame"`, or a `panel_data` with `"panelr"` or `"panel_data"`. `...` sends additional arguments to the functions used to declare those objects.

When using `panel_convert`, be aware that any grouping will be lost, and you must have the relevant package of your `to` option installed (`tsibble`, `plm`, or `panelr`). When your `data` object is a `pdata.frame`, it is recommended to also have `sjlabelled` installed.

All valid objects of the non-`pibble` types can be converted to `pibbles`, but the reverse is not true, since `pibble` does not enforce some strict requirements that other types do:


Feature/Requirement    |  `pibble` |  `tsibble`|  `pdata.frame` |  `panel_data` 
-----------------------|-----------|-----------|----------------|------------
ID                     | `.i`      | `key`     | `index[1]`     | `id` 
Time                   | `.t`      | `index`   | `index[2]`     | `wave` 
Gap control            | `.d`      | `regular` | No             | No  
ID must exist          | No        | No        | Yes            | Yes 
Time must exist        | No        | Yes       | Yes            | Yes[1] 
Only one ID variable[2]| No        | No        | Yes            | Yes 
Unique identification  | No        | Yes       | No[3]          | No[3]

[1] `pdata.frame` does not require that time be provided, but if not provided will create it based on original ordering of the data. The `pdata.frame` option to set `index` equal to an integer for a balanced panel and have it figure out the rest by itself is not supported.

[2] Use `id_variable()` (described below) to generate a single ID variable from multiple if one is required.

[3] `pdata.frame` and `panel_data` do not require that ID and time uniquely identify the observations on declaring the data, but functions in these packages may not work correctly without unique identification.

In addition to the above, be aware that the different packages have different requirements on which variable classes can be Time variables. `time_variable()` (described below) can build an integer variable that will work in all packages.

------

# ID and Time Variables

Any panel data dataset is defined by ID variable(s) (`.i`) that indicate the individual person/firm/country/etc. you're talking about, and a time variable (`.t`) that tell you when the observation in question was recorded.

While `pmdplyr` allows multiple ID variables, many panel data packages only allow one. `id_variable()` allows you to turn multiple ID variables into a single variable.

Many panel data packages, including `pmdplyr`, prefer a time variable that is an integer, so that the difference between each period is 1 (or `.d` in `pmdplyr`'s case). The function `time_variable()` can help you generate an integer time variable from a `Date`-class variable, or from one or more variables containing, for example, year and month.

## id_variable()

`id_variable()` syntax follows:

```{r, eval=FALSE}
id_variable(...,
  .method = "number",
  .minwidth = FALSE
)
```

where `...` is the set of identity variables that you want to combine into a single one (or, potentially, a single variable you'd like to encode numerically).

`.method` describes the way in which you'd like the variable encoded:

* `.method = number` assigns consecutive numeric codes in the original order they appear in the data.
* `.method = random` assigns numeric codes from `0` to `10*N` (where `N` is the number of codes to be assigned) in a random order, so it will be more difficult to uncover the original identity variables.
* `.method = character` preserves all original information and combines the variables together into a string, adding spacing to ensure uniqueness. Set `.minwidth = TRUE` to remove the spacing, although this may lead to non-uniqueness in some cases.

```{r}
df <- data.frame(
  country = c(
    "US", "US", "US", "US",
    "ENG", "ENG", "ENG", "ENG"
  ),
  city = c(
    "NYC", "NYC", "Cambridge", "NYC",
    "Cambridge", "London", "Manchester", "Manchester"
  )
) %>%
  mutate(
    numeric_ID = id_variable(country, city),
    random_ID = id_variable(country, city, .method = "random"),
    char_ID = id_variable(country, city, .method = "character")
  )

df
```

## time_variable()

`time_variable()` syntax follows:

```{r, eval = FALSE}
time_variable(...,
  .method = "present",
  .datepos = NA,
  .start = 1,
  .skip = NA,
  .breaks = NA,
  .turnover = NA,
  .turnover_start = NA
)
```

Where `...` is the set of variables that you want to combine into a single, `integer`-class time variable. The rest of the options determine how the variable(s) will be read or transformed; the need for each varies depending on the structure of the original data and which `.method` is used.

`.method` can take the values:

* `.method = "present"` will assume that, even if each individual may have some missing periods, each period is present in your data *somewhere*, and so simply numbers, in order, all the time periods observed in the data.
* `.method = "year"` can be used with a single `Date`/`POSIX`/etc.-type variable (anything that allows `lubridate::date()`) and will extract the year from it. Or, use it with a character or numeric variable and indicate with `.datepos` the character/digit positions that hold the year in YY or YYYY format.  If combined with `.breaks` or `.skip`, will instead set the earliest year in the data to 1 rather than returning the actual year.
* `.method = "month"` can be used with a single `Date`/`POSIX`/etc.-type variable (anything that allows `lubridate::date()`). It will give the earliest-observed month in the data set a value of `1`, and will increment from there. Or, use it with a character or numeric variable and indicate with `.datepos` the character/digit positions that hold the year and month in YYMM or YYYYMM format (note that if your variable is in MMYYYY format, for example, you can just give a `.datepos` argument like `c(3:6,1:2)`). Months turn over on the `.start` day of the month, which is by default 1.
* `.method = "week"` can be used with a single `Date`/`POSIX`/etc.-type variable (anything that allows `lubridate::date()`). It will give the earliest-observed week in the data set a value of `1`, and will increment from there. Weeks turn over on the `.start` day, which is by default 1 (Monday). Note that this method always starts weeks on the same day of the week, which is different from standard `lubridate` procedure of counting sets of 7 days starting from January 1.
* `.method = "day"` can be used with a single `Date`/`POSIX`/etc.-type variable (anything that allows `lubridate::date()`). It will give the earliest-observed day in the data set a value of `1`, and increment from there. Or, use it with a character or numeric variable and indicate with `.datepos` the character/digit positions that hold the year and month in YYMMDD or YYYYMMDD format. To skip certain days of the week, such as weekends, use the `.skip` option.
* `.method = "turnover"` can be used when you have more than one variable in variable and they are all numeric nonnegative integers. Set the `.turnover` option to indicate the highest value each variable takes before it starts over, and set `.turnover_start` to indicate what value it takes when it starts over. Cannot be combined with `.skip` or `.breaks`. Doesn't work with any variable for which the turnover values change, i.e. it doesn't play well with days-in-month - if you'd like to do something like year-month-day-hour, I recommend running `.method="day"` once with just the year-month-day variable, and then taking the result and combining *that* with hour in `.method = "turnover"`.

```{r}
data(SPrail)

# Since we have a date variable, we can easily create integers that increment for each
# year, or for each month, etc.
# Likely we'd only really need one of these four, depending on our purposes
SPrail <- SPrail %>%
  dplyr::mutate(
    year_time_id = time_variable(insert_date, .method = "year"),
    month_time_id = time_variable(insert_date, .method = "month"),
    week_time_id = time_variable(insert_date, .method = "week"),
    day_time_id = time_variable(insert_date, .method = "day")
  )

# Let's see what we've got
SPrail %>% 
  select(insert_date, ends_with("time_id")) %>% 
  head()

# Perhaps I'd like quarterly data
# (although in this case there are only two months, not much variation there)
SPrail <- SPrail %>%
  dplyr::mutate(quarter_time_id = time_variable(insert_date,
    .method = "month",
    .breaks = c(1, 4, 7, 10)
  ))
# Should line up properly with month
SPrail %>% 
  count(month_time_id, quarter_time_id)

# Maybe I'd like Monday to come immediately after Friday!
SPrail <- SPrail %>%
  dplyr::mutate(weekday_time_id = time_variable(insert_date,
    .method = "day",
    .skip = c(6, 7)
  ))

# Perhaps I'm interested in ANY time period in the data and just want to enumerate them in order
SPrail <- SPrail %>%
  dplyr::mutate(any_present_time_id = time_variable(insert_date,
    .method = "present"
  ))

# Note the weekday_time_id NAs - these are weekends! We told it to skip those.
head(SPrail %>% select(insert_date, day_time_id, weekday_time_id, any_present_time_id))

# Maybe instead of being given a nice time variable, I was given it in string form
SPrail <- SPrail %>% dplyr::mutate(time_string = as.character(insert_date))
# As long as the character positions are consistent we can still use it
SPrail <- SPrail %>%
  dplyr::mutate(day_from_string_id = time_variable(time_string,
    .method = "day",
    .datepos = c(3, 4, 6, 7, 9, 10)
  ))
# Results are identical from using the actual Date variable
cor(SPrail$day_time_id, SPrail$day_from_string_id)
```
